---
title: "Hands-on Exercise 7a"
subtitle: "Modelling, Visualising and Analysing Network Data with R"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Overview

This hands-on exercise recreates the key techniques from Chapter 27 of *R for Visual Analytics*. Using the GAStech email corpus, it models the organisation as a **graph** and visualises it with `tidygraph` + `ggraph` (static) and `visNetwork` (interactive), including centrality-based node sizing.

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(
  igraph, tidygraph, ggraph, visNetwork,
  lubridate, graphlayouts, ggforce, tidyverse
)

GAStech_nodes <- read_csv("data/GAStech_email_node.csv")
GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")
```

## Data Wrangling

The send date is parsed and the weekday derived. Edges are then aggregated to weighted, work-related links between different people.

```{r}
#| message: false
#| warning: false

GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SendDate, label = TRUE, abbr = FALSE))

GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
  summarise(Weight = n(), .groups = "drop") %>%
  filter(source != target) %>%
  filter(Weight > 1)
```

# 1. Building the Graph Object

`tbl_graph()` combines the node and edge tables into a `tidygraph` object.

```{r}
GAStech_graph <- tbl_graph(
  nodes = GAStech_nodes,
  edges = GAStech_edges_aggregated,
  directed = TRUE
)

GAStech_graph
```

# 2. Static Network with ggraph

## 2.1 Basic Plot

```{r}
#| fig-width: 9
#| fig-height: 7
ggraph(GAStech_graph, layout = "nicely") +
  geom_edge_link(aes(width = Weight), alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), size = 3) +
  theme_graph()
```

## 2.2 Faceting by Weekday

```{r}
#| fig-width: 11
#| fig-height: 7
ggraph(GAStech_graph, layout = "nicely") +
  geom_edge_link(aes(width = Weight), alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), size = 2) +
  facet_edges(~ Weekday) +
  theme_graph(foreground = "grey80", fg_text_colour = "white")
```

## 2.3 Sizing Nodes by Betweenness Centrality

Betweenness centrality highlights the people who act as bridges in the communication network.

```{r}
#| fig-width: 9
#| fig-height: 7
GAStech_graph <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness())

ggraph(GAStech_graph, layout = "fr") +
  geom_edge_link(aes(width = Weight), alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department, size = betweenness_centrality)) +
  theme_graph()
```

## 2.4 Detecting Communities

`tidygraph` inherits `igraph`'s community-detection algorithms via its `group_*` functions. Here `group_edge_betweenness()` partitions the graph into communities, which are then mapped to node colour.

```{r}
#| fig-width: 9
#| fig-height: 7
#| message: false
#| warning: false
g <- GAStech_graph %>%
  mutate(community = as.factor(
    group_edge_betweenness(
      weights = Weight,
      directed = TRUE))) %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(width = Weight), alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = community))

g + theme_graph()
```

To make communities easier to read, `geom_mark_hull()` (from `ggforce`) draws a hull around each community found by `group_optimal()`, while nodes are still coloured by `Department` and sized by betweenness centrality.

```{r}
#| fig-width: 9
#| fig-height: 7
#| message: false
#| warning: false
g <- GAStech_graph %>%
  activate(nodes) %>%
  mutate(community = as.factor(
    group_optimal(weights = Weight)),
    betweenness_measure = centrality_betweenness()) %>%
  ggraph(layout = "fr") +
  geom_mark_hull(
    aes(x, y,
        group = community,
        fill = community),
    alpha = 0.2,
    expand = unit(0.3, "cm"),
    radius = unit(0.3, "cm")
  ) +
  geom_edge_link(aes(width = Weight), alpha = 0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(fill = Department,
                      size = betweenness_measure),
                  color = "black",
                  shape = 21)

g + theme_graph()
```

# 3. Interactive Network with visNetwork

`visNetwork` expects the columns `from` and `to`; the aggregated edges are renamed accordingly.

```{r}
GAStech_edges_vis <- GAStech_edges_aggregated %>%
  rename(from = source, to = target)

visNetwork(GAStech_nodes, GAStech_edges_vis) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(
    highlightNearest = list(enabled = TRUE, degree = 1, hover = TRUE),
    nodesIdSelection = TRUE
  ) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

# Learning Reflection

In this exercise, I learned that network data needs two tables - nodes and edges - and that most of the analytical work happens before plotting: parsing dates, filtering to meaningful (work-related) links, and aggregating them into weighted edges. `tidygraph` lets me treat the graph like a tidy data frame, so computing **betweenness centrality** and mapping it to node size is a single `mutate()`. `ggraph` produces publication-quality static layouts (and can facet by weekday), while `visNetwork` makes the same graph interactive for exploration.
