---
title: "Hands-on Exercise 2a"
subtitle: "Beyond ggplot2 Fundamentals"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Overview

This hands-on exercise recreates the key techniques from Chapter 2 of *R for Visual Analytics*. The focus is on extending `ggplot2` with three companion packages: `ggrepel` for readable annotations, `ggthemes` and `hrbrthemes` for professional themes, and `patchwork` for composing multiple plots into a single figure.

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(
  ggrepel,
  patchwork,
  ggthemes,
  hrbrthemes,
  tidyverse
)

exam_data <- read_csv("Exam_data.csv")
```

## Data Preview

```{r}
DT::datatable(
  exam_data,
  class = "compact",
  options = list(
    pageLength = 10,
    scrollX = TRUE
  )
)
```

# 1. Annotation with ggrepel

## 1.1 The Overlapping-label Problem

Using `geom_label()` directly causes labels to overlap and become unreadable.

```{r}
ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm, linewidth = 0.5) +
  geom_label(aes(label = ID), hjust = 0.5, vjust = -0.5) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## 1.2 Repelling Labels with `geom_label_repel()`

`ggrepel` replaces the geoms with repel-aware versions that push labels apart.

```{r}
ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm, linewidth = 0.5) +
  geom_label_repel(aes(label = ID), fontface = "bold") +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

# 2. Beyond ggplot2 Themes

## 2.1 Built-in Theme

```{r}
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, boundary = 100, color = "grey25", fill = "grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores")
```

## 2.2 The ggthemes Package

`ggthemes` provides themes that mimic well-known publications, e.g. *The Economist*.

```{r}
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, boundary = 100, color = "grey25", fill = "grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```

## 2.3 The hrbrthemes Package

`hrbrthemes` focuses on typographic detail and axis label placement with `theme_ipsum()`.

```{r}
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, boundary = 100, color = "grey25", fill = "grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18, base_size = 15, grid = "Y")
```

# 3. Composing Figures with patchwork

## 3.1 Creating the Individual Plots

```{r}
p1 <- ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, boundary = 100, color = "grey25", fill = "grey90") +
  coord_cartesian(xlim = c(0, 100)) +
  ggtitle("Distribution of Maths scores")

p2 <- ggplot(exam_data, aes(x = ENGLISH)) +
  geom_histogram(bins = 20, boundary = 100, color = "grey25", fill = "grey90") +
  coord_cartesian(xlim = c(0, 100)) +
  ggtitle("Distribution of English scores")

p3 <- ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm, linewidth = 0.5) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
  ggtitle("English scores versus Maths scores")
```

## 3.2 Two Plots Side by Side with `+`

```{r}
p1 + p2
```

## 3.3 Complex Layout with `|` and `/`

Two plots on top, one spanning the bottom.

```{r}
(p1 / p2) | p3
```

## 3.4 Auto-tagging Sub-plots

```{r}
((p1 / p2) | p3) +
  plot_annotation(tag_levels = "I")
```

## 3.5 Inset Plots with `inset_element()`

```{r}
p3 + inset_element(p2, left = 0.02, bottom = 0.7, right = 0.5, top = 1)
```

## 3.6 Combining patchwork and ggtheme

```{r}
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```

# Learning Reflection

In this exercise, I learned that `ggplot2` is only the starting point. `ggrepel` solves the very common problem of overlapping text annotations, turning an unreadable labelled scatterplot into a clear one. `ggthemes` and `hrbrthemes` let me adopt a polished, publication-grade look without hand-tuning every theme element. `patchwork` is the most powerful of the three for reporting: with simple `+`, `|`, and `/` operators I can assemble several plots into one coherent figure, tag the panels, and even inset one plot inside another.
