---
title: "Hands-on Exercise 10a"
subtitle: "Information Dashboard Design: R methods"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Overview

This hands-on exercise recreates the key techniques from Chapter 31 of *R for Visual Analytics*. It builds an information dashboard for the CoffeeChain sales data: a **bullet chart** and **sparklines** in ggplot2, a **static dashboard** with `gt` and `gtExtras`, and an **interactive dashboard** with `reactable` and `reactablefmtr`.

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(lubridate, ggthemes, reactable, reactablefmtr, gt, gtExtras, tidyverse)

coffeechain <- read_rds("data/rds/CoffeeChain.rds")
```

## Data Preparation

Sales and budgeted sales are aggregated to the product level.

```{r}
product <- coffeechain %>%
  group_by(Product) %>%
  summarise(
    target = sum(`Budget Sales`),
    current = sum(Sales)
  ) %>%
  ungroup()
```

# 1. Bullet Chart in ggplot2

The grey columns give the qualitative ranges, the black bar is the actual value and the red line marks the target.

```{r}
ggplot(product, aes(Product, current)) +
  geom_col(aes(Product, max(target) * 1.01), fill = "grey85", width = 0.85) +
  geom_col(aes(Product, target * 0.75), fill = "grey60", width = 0.85) +
  geom_col(aes(Product, target * 0.5), fill = "grey50", width = 0.85) +
  geom_col(aes(Product, current), width = 0.35, fill = "black") +
  geom_errorbar(aes(y = target, x = Product, ymin = target, ymax = target),
                width = 0.4, colour = "red", linewidth = 1) +
  coord_flip()
```

# 2. Sparklines in ggplot2

## 2.1 Preparing the Data

```{r}
#| message: false

sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)

mins <- group_by(sales_report, Product) %>% slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>% slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>% filter(Month == max(Month))

quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(
    quart1 = quantile(Sales, 0.25),
    quart2 = quantile(Sales, 0.75)
  ) %>%
  right_join(sales_report)
```

## 2.2 The Sparklines

```{r}
#| fig-width: 14
#| fig-height: 12
#| message: false

ggplot(sales_report, aes(x = Month, y = Sales)) +
  facet_grid(Product ~ ., scales = "free_y") +
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2), fill = "grey90") +
  geom_line(linewidth = 0.3) +
  geom_point(data = mins, col = "red") +
  geom_point(data = maxs, col = "blue") +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) +
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title = element_blank(), axis.text.y = element_blank(),
        axis.ticks = element_blank(), strip.text = element_blank())
```

# 3. Static Dashboard with gt and gtExtras

## 3.1 Bullet Chart with `gt_plt_bullet()`

```{r}
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current, target = target, width = 60,
                palette = c("lightblue", "black")) %>%
  gt_theme_538()
```

## 3.2 Sparklines with `gt_plt_sparkline()`

The monthly sales are converted into a **list column**, which is what the gtExtras plotting functions require.

```{r}
#| message: false

report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate(Month = month(Date, label = TRUE, abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()

report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales), .groups = "drop") %>%
  gt() %>%
  gt_plt_sparkline(`Monthly Sales`, same_limit = FALSE)
```

# 4. Interactive Dashboard with reactablefmtr

`react_sparkline()` renders the same sparklines as an interactive `reactable` table, here highlighting the minimum and maximum months and labelling the first and last values.

```{r}
report_list <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales), .groups = "drop")

reactable(
  report_list,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report_list,
        highlight_points = highlight_points(min = "red", max = "blue"),
        labels = c("first", "last")
      )
    )
  )
)
```

# Learning Reflection

In this exercise, I learned that an information dashboard is built from small, dense, comparison-first components. A **bullet chart** compresses an actual value, a target and qualitative ranges into one bar; a **sparkline** compresses a whole monthly time series into a tiny line. Building these in ggplot2 shows the underlying grammar (layered `geom_col`, `facet_grid`, `geom_ribbon` for the quartile band), while `gt` + `gtExtras` (`gt_plt_bullet`, `gt_plt_sparkline`) embed the same graphics directly inside a table for a compact, scannable report.
