---
title: "Hands-on Exercise 5c"
subtitle: "Heatmap for Visualising and Analysing Multivariate Data"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Overview

This hands-on exercise recreates the key techniques from Chapter 14 of *R for Visual Analytics*. A **heatmap** encodes each value as a colour, letting us read a whole multivariate table at a glance. Using the 2018 World Happiness data, this exercise builds both a static clustered heatmap and an interactive one with `heatmaply`.

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(seriation, dendextend, heatmaply, tidyverse)

wh <- read_csv("data/WHData-2018.csv")
```

## Data Preparation

The country column is promoted to row names and the data frame is converted to a numeric matrix, which is the input every heatmap function expects.

```{r}
wh <- as.data.frame(wh)
row.names(wh) <- wh$Country
wh_matrix <- data.matrix(wh)
```

# 1. Static Heatmap with `heatmap()`

## 1.1 Without Clustering

```{r}
#| fig-width: 8
#| fig-height: 9
heatmap(
  wh_matrix,
  Rowv = NA,
  Colv = NA
)
```

## 1.2 With Clustering and Column Scaling

Scaling each column to a common range prevents the large-magnitude variables from dominating the colour scale, and the dendrograms reorder rows and columns by similarity.

```{r}
#| fig-width: 8
#| fig-height: 9
heatmap(
  wh_matrix,
  scale = "column",
  cexRow = 0.6,
  cexCol = 0.8,
  margins = c(10, 4)
)
```

# 2. Interactive Heatmap with heatmaply

## 2.1 Normalising the Variables

The whisker and country/region columns are dropped, and `normalize()` rescales every remaining variable to 0-1 so they are comparable.

```{r}
#| fig-width: 8
#| fig-height: 9
heatmaply(
  normalize(wh_matrix[, -c(1, 2, 4, 5)]),
  Colv = NA,
  seriate = "none",
  colors = Blues,
  k_row = 5,
  margins = c(NA, 200, 60, NA),
  fontsize_row = 4,
  fontsize_col = 5,
  main = "World Happiness 2018 (normalised)",
  xlab = "Indicators",
  ylab = "Countries"
)
```

## 2.2 Reordering with Seriation

Optimal-leaf-ordering ("OLO") rotates the dendrogram branches to minimise the distance between adjacent rows, revealing clearer blocks.

```{r}
#| fig-width: 8
#| fig-height: 9
heatmaply(
  normalize(wh_matrix[, -c(1, 2, 4, 5)]),
  seriate = "OLO",
  k_row = 5,
  margins = c(NA, 200, 60, NA),
  fontsize_row = 4,
  fontsize_col = 5,
  main = "World Happiness 2018 (seriated)"
)
```

# Learning Reflection

In this exercise, I learned that a heatmap turns an overwhelming numeric table into a colour field the eye can scan instantly. The two essential pre-processing steps are getting the data into a numeric **matrix** with meaningful row names, and **normalising** the columns so no single indicator dominates the colour scale. The static `heatmap()` is quick, but `heatmaply` adds interactivity, dendrogram control, and seriation (`OLO`) that reorders the rows into clearly separated clusters of similarly-happy countries.
