---
title: "Hands-on Exercise 5b"
subtitle: "Building Ternary Plot with R"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Overview

This hands-on exercise recreates the key techniques from Chapter 13 of *R for Visual Analytics*. A **ternary plot** shows the composition of three mutually exclusive parts that sum to a whole. Here it is used to display the age structure (young / economically active / old) of Singapore's planning subzones, using `ggtern` for the static plot and `plotly` for the interactive version.

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(plotly, ggtern, tidyverse)

pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv")
```

## Data Preparation

The tidy data is reshaped so each subzone becomes one row, then the age groups are aggregated into three classes: **YOUNG** (0-24), **ACTIVE** (25-64) and **OLD** (65+). Only the 2018 records with a positive population are kept.

```{r}
#| message: false
#| warning: false

agpop_mutated <- pop_data %>%
  mutate(Year = as.character(Year)) %>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8])) %>%
  mutate(ACTIVE = rowSums(.[9:16])) %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018) %>%
  filter(TOTAL > 0)
```

```{r}
DT::datatable(
  agpop_mutated %>% select(PA, SZ, YOUNG, ACTIVE, OLD, TOTAL),
  class = "compact",
  options = list(pageLength = 10, scrollX = TRUE)
)
```

# 1. Static Ternary Plot with ggtern

## 1.1 Basic Ternary Plot

```{r}
ggtern(data = agpop_mutated, aes(x = YOUNG, y = ACTIVE, z = OLD)) +
  geom_point()
```

## 1.2 With the `theme_rgbw()` Theme

```{r}
ggtern(data = agpop_mutated, aes(x = YOUNG, y = ACTIVE, z = OLD)) +
  geom_point() +
  labs(title = "Population structure of Singapore subzones, 2018") +
  theme_rgbw()
```

# 2. Interactive Ternary Plot with plotly

```{r}
# reusable axis-label formatter
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"),
  baxis = axis("Active"),
  caxis = axis("Old")
)

plot_ly(
  agpop_mutated,
  a = ~YOUNG,
  b = ~ACTIVE,
  c = ~OLD,
  color = I("#1f77b4"),
  type = "scatterternary",
  text = ~paste0("Subzone: ", SZ)
) %>%
  layout(
    title = "Population structure, 2018",
    ternary = ternaryAxes
  )
```

# Learning Reflection

In this exercise, I learned that a ternary plot is the right tool when three parts must sum to a whole - here, the young / active / old split of each subzone. The hardest part was the data preparation: reshaping the tidy table with `spread()` and aggregating the five-year age bands into three groups before anything could be plotted. `ggtern` produces a clean static composition plot, while `plotly`'s `scatterternary` type adds hover tooltips so I can identify exactly which subzone each point represents.
