---
title: "Hands-on Exercise 9c"
subtitle: "Analytical Mapping"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Overview

This hands-on exercise recreates the key techniques from Chapter 23 of *R for Visual Analytics*. Using a prepared `sf` object of Nigeria's water points at the LGA level, it demonstrates **analytical mapping** - why we map *rates* rather than raw *counts*, and how to build a **percentile map** to expose extreme values.

> **Note:** `sf` and `tmap` depend on the system libraries GDAL, GEOS and PROJ.

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(sf, tmap, tidyverse)

NGA_wp <- read_rds("data/rds/NGA_wp.rds")
```

## Deriving Rates

The count of functional water points is turned into a **proportion** of all water points in each LGA.

```{r}
NGA_wp <- NGA_wp %>%
  mutate(
    pct_functional = wp_functional / total_wp,
    pct_nonfunctional = wp_nonfunctional / total_wp
  )
```

# 1. Why Rates, Not Counts

## 1.1 Mapping Raw Counts

A count map mostly shows where there are *many* water points - not where they are *working*.

```{r}
#| fig-width: 9
#| fig-height: 6
tmap_mode("plot")

tm_shape(NGA_wp) +
  tm_polygons(
    fill = "wp_functional",
    fill.scale = tm_scale_intervals(n = 6, style = "quantile", values = "brewer.blues"),
    col = "white",
    lwd = 0.1
  ) +
  tm_title("Count of functional water points")
```

## 1.2 Mapping the Rate

The rate map answers the real question - the *share* of water points that are functional.

```{r}
#| fig-width: 9
#| fig-height: 6
tm_shape(NGA_wp) +
  tm_polygons(
    fill = "pct_functional",
    fill.scale = tm_scale_intervals(n = 6, style = "quantile", values = "brewer.blues"),
    col = "white",
    lwd = 0.1
  ) +
  tm_title("Percentage of functional water points")
```

# 2. Percentile Map

A percentile map is a special choropleth with six unequal classes (1%, 10%, 50%, 90%, 99%) that spotlights the extreme low and high LGAs.

```{r}
percent <- c(0, 0.01, 0.1, 0.5, 0.9, 0.99, 1)

var <- NGA_wp %>%
  st_set_geometry(NULL) %>%
  pull(pct_functional)

# dedupe breaks: ties (many LGAs at 0% or 100%) can make adjacent quantiles equal
breaks <- unique(quantile(var, percent, na.rm = TRUE))
```

```{r}
#| fig-width: 9
#| fig-height: 6
tm_shape(NGA_wp) +
  tm_polygons(
    fill = "pct_functional",
    fill.scale = tm_scale_intervals(breaks = breaks, values = "-brewer.rd_bu"),
    fill.legend = tm_legend(title = "Percentile"),
    col = "white",
    lwd = 0.1
  ) +
  tm_title("Percentile map: functional water point rate")
```

# 3. Box Map

A box map extends the boxplot idea onto the map: it splits the LGAs into six classes - a lower-outlier class, the four inter-quartile classes, and an upper-outlier class - so that both the typical spread and the extreme values are visible at once.

## 3.1 Helper Functions

`boxbreaks()` computes the seven break points (fences and quartiles), and `get.var()` pulls a variable out of the `sf` object as a plain numeric vector.

```{r}
boxbreaks <- function(v, mult = 1.5) {
  qv <- unname(quantile(v, na.rm = TRUE))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode = "numeric", length = 7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) {  # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}

get.var <- function(vname, df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[, 1])
  return(v)
}
```

## 3.2 The Box Map Function

`boxmap()` wires the helpers into a tmap v4 choropleth, labelling the six classes from lower outlier through to upper outlier.

```{r}
boxmap <- function(vnam, df,
                   legtitle = "Box map class",
                   mtitle = "Box Map") {
  var <- get.var(vnam, df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons(
      fill = vnam,
      fill.scale = tm_scale_intervals(
        breaks = bb,
        values = "-brewer.rd_bu",
        labels = c("lower outlier", "< 25%", "25% - 50%",
                   "50% - 75%", "> 75%", "upper outlier")
      ),
      fill.legend = tm_legend(title = legtitle),
      col = "white",
      lwd = 0.1
    ) +
    tm_title(mtitle)
}
```

## 3.3 Box Map of the Functional Rate

Applying `boxmap()` to `pct_functional` highlights the outlier LGAs at both ends of the distribution.

```{r}
#| fig-width: 9
#| fig-height: 6
tmap_mode("plot")

boxmap("pct_functional", NGA_wp)
```

# Learning Reflection

In this exercise, I learned the single most important lesson in thematic mapping: **map rates, not counts**. A count map of functional water points simply re-draws where water points are dense, whereas the rate map reveals where provision is actually failing. I also learned to build a **percentile map** using custom `quantile()` breaks, which deliberately isolates the extreme 1% and 99% LGAs - the places that most need attention - instead of hiding them inside broad equal classes.
