---
title: "Hands-on Exercise 6a"
subtitle: "Designing Web-enabled Visual Analytics Applications with Shiny"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Overview

This hands-on exercise accompanies Lesson 6 of ISSS608. Unlike the other hands-on exercises, Shiny has **no R for Visual Analytics book chapter** - it is taught through the lesson slides. This exercise therefore builds a small **interactive Shiny web application**: an exam-scores explorer on the `Exam_data.csv` dataset, deployed to shinyapps.io.

The app demonstrates the three Shiny building blocks: a **UI** (`page_sidebar` with inputs), **reactivity** (outputs that update when inputs change), and **deployment** to a public server.

## Live Application

The deployed app is embedded below. It lets you pick a subject, choose a plot type (histogram / boxplot / density), group by gender, race or class, and read the underlying records.

```{=html}
<iframe src="https://jedsjed.shinyapps.io/isss608-hoe6a/"
        width="100%" height="760"
        style="border:1px solid #e0e0e0; border-radius:8px;"
        loading="lazy">
</iframe>
```

::: {.callout-note}
If the frame is blank, the free shinyapps.io instance may be asleep - open it directly at
[jedsjed.shinyapps.io/isss608-hoe6a](https://jedsjed.shinyapps.io/isss608-hoe6a/){target="_blank"}.
:::

## The Application Code

The entire app is a single `app.R`. The **UI** describes the layout and inputs; the **server** defines reactive expressions and renders the outputs.

```{r}
#| eval: false
#| code-fold: show

library(shiny)
library(bslib)
library(tidyverse)
library(plotly)
library(DT)

exam <- readr::read_csv("Exam_data.csv", show_col_types = FALSE)

subjects <- c("MATHS", "ENGLISH", "SCIENCE")
genders  <- sort(unique(exam$GENDER))
classes  <- sort(unique(exam$CLASS))

ui <- page_sidebar(
  title = "ISSS608 · Hands-on 6a: Exam Scores Explorer",
  theme = bs_theme(version = 5, bootswatch = "cosmo", primary = "#1e3a6b"),
  sidebar = sidebar(
    title = "Controls",
    selectInput("subject", "Subject", choices = subjects, selected = "MATHS"),
    radioButtons("plot_type", "Plot type",
                 choices = c("Histogram", "Boxplot", "Density")),
    selectInput("group", "Colour by",
                choices = c("None", "GENDER", "RACE", "CLASS"), selected = "GENDER"),
    conditionalPanel(
      "input.plot_type == 'Histogram'",
      sliderInput("bins", "Number of bins", min = 5, max = 40, value = 20)
    ),
    hr(), strong("Filter the data"),
    checkboxGroupInput("f_gender", "Gender", choices = genders, selected = genders, inline = TRUE),
    selectizeInput("f_class", "Class", choices = classes, selected = classes, multiple = TRUE)
  ),
  layout_columns(
    col_widths = c(6, 6),
    value_box("Students in view", textOutput("n_students"),
              showcase = bsicons::bs_icon("people")),
    value_box("Mean score", textOutput("mean_score"),
              showcase = bsicons::bs_icon("bar-chart-line"))
  ),
  card(card_header("Distribution"), plotlyOutput("dist_plot", height = "380px")),
  card(card_header("Records"), DTOutput("tbl"))
)

server <- function(input, output, session) {
  # the Gender / Class controls actually subset the data
  filtered <- reactive({
    d <- exam
    if (length(input$f_gender)) d <- dplyr::filter(d, GENDER %in% input$f_gender)
    if (length(input$f_class))  d <- dplyr::filter(d, CLASS %in% input$f_class)
    d
  })
  output$n_students <- renderText(nrow(filtered()))
  output$mean_score <- renderText(round(mean(filtered()[[input$subject]], na.rm = TRUE), 1))
  # ... renderPlotly() switches geom by input$plot_type and colours by input$group ...
  output$tbl <- renderDT(datatable(filtered(), options = list(pageLength = 8)))
}

shinyApp(ui, server)
```

The full source lives in the `shiny_ex6a/` folder of this repository.

## Deploying to shinyapps.io

```{r}
#| eval: false
#| code-fold: show

library(rsconnect)
# one-time: rsconnect::setAccountInfo(name = "...", token = "...", secret = "...")
deployApp(appDir = "shiny_ex6a", appName = "isss608-hoe6a")
```

# Learning Reflection

In this exercise, I learned that a Shiny app is built from three moving parts. The **UI** is just a nested layout of input and output placeholders - `bslib`'s `page_sidebar`, `card` and `value_box` make it look professional with almost no CSS. **Reactivity** is the heart of Shiny: an output like `renderPlotly()` automatically re-runs whenever an input it reads (`input$subject`, `input$plot_type`, `input$group`) changes, so I never write update logic by hand. Finally, `rsconnect::deployApp()` pushes the whole thing - code and data - to shinyapps.io, turning a local script into a public web application. This is the foundation for the group project's larger dashboard.
