---
title: "Hands-on Exercise 8a"
subtitle: "Visualising and Analysing Time-oriented Data"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Overview

This hands-on exercise recreates the key techniques from Chapter 17 of *R for Visual Analytics*. It covers three time-oriented visualisation methods: a **calendar heatmap** of cyber-attack activity, a **cycle plot** of visitor arrivals, and a **slopegraph** of rice yields.

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(
  data.table, scales, viridis, lubridate, ggthemes,
  gridExtra, readxl, knitr, CGPfunctions, tidyverse
)
```

# 1. Calendar Heatmap

## 1.1 Importing and Deriving Weekday / Hour

```{r}
#| message: false
#| warning: false

attacks <- read_csv("data/eventlog.csv")

make_hr_wkday <- function(ts, sc, tz) {
  real_times <- ymd_hms(ts, tz = tz[1], quiet = TRUE)
  dt <- data.table(
    source_country = sc,
    wkday = weekdays(real_times),
    hour  = hour(real_times)
  )
  return(dt)
}

wkday_levels <- c("Saturday", "Friday", "Thursday",
                  "Wednesday", "Tuesday", "Monday", "Sunday")

attacks <- attacks %>%
  group_by(tz) %>%
  do(make_hr_wkday(.$timestamp, .$source_country, .$tz)) %>%
  ungroup() %>%
  mutate(
    wkday = factor(wkday, levels = wkday_levels),
    hour  = factor(hour, levels = 0:23)
  )
```

## 1.2 Building the Heatmap

```{r}
#| fig-width: 10
#| fig-height: 4

grouped <- attacks %>%
  count(wkday, hour) %>%
  ungroup() %>%
  na.omit()

ggplot(grouped, aes(x = hour, y = wkday, fill = n)) +
  geom_tile(color = "white", linewidth = 0.1) +
  theme_tufte(base_family = "Helvetica") +
  coord_equal() +
  scale_fill_gradient(name = "# of attacks", low = "sky blue", high = "dark blue") +
  labs(x = NULL, y = NULL, title = "Attacks by weekday and time of day") +
  theme(
    axis.ticks = element_blank(),
    plot.title = element_text(hjust = 0.5),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6)
  )
```

# 2. Cycle Plot

A cycle plot places each month's yearly trend side by side, with a red mean line, so both the seasonal pattern and the year-on-year trend are visible at once.

## 2.1 Data Preparation

```{r}
#| message: false
#| warning: false

air <- read_xlsx("data/arrivals_by_air.xlsx")

air$month <- factor(month(air$`Month-Year`),
                    levels = 1:12, labels = month.abb, ordered = TRUE)
air$year <- year(ymd(air$`Month-Year`))

Japan <- air %>%
  select(Japan, month, year) %>%
  filter(year >= 2010)

hline.data <- Japan %>%
  group_by(month) %>%
  summarise(avgvalue = mean(Japan), .groups = "drop")
```

## 2.2 Building the Cycle Plot

```{r}
#| fig-width: 11
#| fig-height: 5

ggplot() +
  geom_line(data = Japan, aes(x = year, y = Japan, group = month), colour = "black") +
  geom_hline(aes(yintercept = avgvalue), data = hline.data,
             linetype = 6, colour = "red", linewidth = 0.5) +
  facet_grid(~ month) +
  labs(
    x = NULL, y = "No. of Visitors",
    title = "Visitor arrivals from Japan by air, cycle plot"
  ) +
  theme_tufte(base_family = "Helvetica")
```

# 3. Slopegraph

A slopegraph compares two time points and highlights which series rose or fell between them.

```{r}
#| fig-width: 8
#| fig-height: 9

rice <- read_csv("data/rice.csv", show_col_types = FALSE)

rice %>%
  mutate(Year = factor(Year)) %>%
  filter(Year %in% c(1961, 1980)) %>%
  newggslopegraph(
    Year, Yield, Country,
    Title = "Rice Yield of Top 11 Asian Countries",
    SubTitle = "1961-1980",
    Caption = "Source: R for Visual Analytics, Chapter 17"
  )
```

# Learning Reflection

In this exercise, I learned three complementary ways to look at time. A **calendar heatmap** exposes recurring daily/weekly rhythms - cyber-attacks cluster at particular hours and weekdays. A **cycle plot** separates the seasonal shape from the long-run trend by laying each month's mini-timeline side by side against its mean. A **slopegraph** strips a change down to its essentials - two points and the direction of the line between them - making winners and losers obvious. The common thread is that each design deliberately reshapes time to answer a specific question.
