---
title: "Hands-on Exercise 9b"
subtitle: "Visualising Geospatial Point Data"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Overview

This hands-on exercise recreates the key techniques from Chapter 22 of *R for Visual Analytics*. It turns an **aspatial** table of Singapore Pools outlets - with X/Y coordinates in the SVY21 projected system - into a geospatial point layer, then draws proportional-symbol maps with `tmap`.

> **Note:** `sf` and `tmap` depend on the system libraries GDAL, GEOS and PROJ.

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(sf, tmap, tidyverse)

sgpools <- read_csv("data/aspatial/SGPools_svy21.csv")
```

## Creating an sf Point Layer

`st_as_sf()` converts the XCOORD/YCOORD columns into geometry. EPSG **3414** is Singapore's SVY21 projected coordinate system.

```{r}
sgpools_sf <- st_as_sf(
  sgpools,
  coords = c("XCOORD", "YCOORD"),
  crs = 3414
)

sgpools_sf
```

# 1. Interactive Proportional Symbol Map

Bubble size encodes total winnings; colour encodes outlet type. `view` mode renders an interactive leaflet map.

```{r}
#| message: false
#| warning: false
tmap_mode("view")

tm_shape(sgpools_sf) +
  tm_bubbles(
    fill = "OUTLET TYPE",
    size = "Gp1Gp2 Winnings",
    col = "black",
    lwd = 1
  )
```

# 2. Faceted Maps by Outlet Type

Small multiples make it easy to compare the spatial spread of the two outlet types.

```{r}
#| fig-width: 10
#| fig-height: 6
tmap_mode("plot")

tm_shape(sgpools_sf) +
  tm_bubbles(
    fill = "OUTLET TYPE",
    size = "Gp1Gp2 Winnings",
    col = "black",
    lwd = 1
  ) +
  tm_facets(by = "OUTLET TYPE", nrow = 1, sync = TRUE)
```

```{r}
#| echo: false
tmap_mode("plot")
```

# Learning Reflection

In this exercise, I learned how to promote an ordinary table into a mappable point layer: the key step is `st_as_sf()` with the correct coordinate columns and the correct **CRS** (SVY21 / EPSG:3414 here). Getting the CRS wrong would place every outlet in the wrong location. Once the layer exists, `tmap` makes proportional-symbol mapping straightforward, and switching to `view` mode turns the same code into an interactive leaflet map - while `tm_facets()` lets me compare outlet types as small multiples.
