---
title: "Hands-on Exercise 3"
subtitle: "Programming Interactive Data Visualisation with R"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Overview

This hands-on exercise recreates the key interactive visualisation techniques from Chapter 3 of *R for Visual Analytics*. The focus is on using `ggiraph`, `plotly`, `DT`, and `crosstalk` to make static R graphics interactive.

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(
  ggiraph,
  plotly,
  patchwork,
  DT,
  tidyverse,
  crosstalk
)

exam_data <- read_csv("Exam_data.csv")
```

## Data Preview

```{r}
DT::datatable(
  exam_data,
  class = "compact",
  options = list(
    pageLength = 10,
    scrollX = TRUE
  )
)
```

# 1. Interactive Data Visualisation with ggiraph

## 1.1 Tooltip Showing Student ID

```{r}
p <- ggplot(exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot"
  ) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(
    title = "Distribution of Maths Scores",
    x = "Maths Score"
  )

girafe(
  ggobj = p,
  width_svg = 7,
  height_svg = 4
)
```

## 1.2 Tooltip Showing Multiple Information

```{r}
exam_data <- exam_data %>%
  mutate(
    tooltip_text = paste0(
      "ID: ", ID,
      "\nClass: ", CLASS,
      "\nGender: ", GENDER,
      "\nRace: ", RACE,
      "\nMaths: ", MATHS
    )
  )

p <- ggplot(exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = tooltip_text),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot"
  ) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(
    title = "Maths Score Distribution with Detailed Tooltip",
    x = "Maths Score"
  )

girafe(
  ggobj = p,
  width_svg = 8,
  height_svg = 4
)
```

## 1.3 Custom Tooltip Style

```{r}
tooltip_css <- "
  background-color: white;
  color: black;
  font-weight: bold;
  padding: 6px;
  border-radius: 4px;
  border: 1px solid #333333;
"

p <- ggplot(exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = tooltip_text),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot"
  ) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(
    title = "Custom Tooltip Style",
    x = "Maths Score"
  )

girafe(
  ggobj = p,
  width_svg = 8,
  height_svg = 4,
  options = list(
    opts_tooltip(css = tooltip_css)
  )
)
```

## 1.4 Hover Effect by Class

```{r}
p <- ggplot(exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot"
  ) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(
    title = "Hover Highlight by Class",
    x = "Maths Score"
  )

girafe(
  ggobj = p,
  width_svg = 8,
  height_svg = 4
)
```

## 1.5 Styled Hover Effect

```{r}
p <- ggplot(exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot"
  ) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(
    title = "Styled Hover Highlight by Class",
    x = "Maths Score"
  )

girafe(
  ggobj = p,
  width_svg = 8,
  height_svg = 4,
  options = list(
    opts_hover(css = "fill: #202020;"),
    opts_hover_inv(css = "opacity:0.2;")
  )
)
```

## 1.6 Tooltip and Hover Combined

```{r}
p <- ggplot(exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(
      tooltip = tooltip_text,
      data_id = CLASS
    ),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot"
  ) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(
    title = "Tooltip and Hover Combined",
    x = "Maths Score"
  )

girafe(
  ggobj = p,
  width_svg = 8,
  height_svg = 4,
  options = list(
    opts_tooltip(css = tooltip_css),
    opts_hover(css = "fill: #202020;"),
    opts_hover_inv(css = "opacity:0.2;")
  )
)
```

## 1.7 Coordinated Multiple Views with ggiraph

```{r}
p1 <- ggplot(exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = ID, tooltip = tooltip_text),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot"
  ) +
  coord_cartesian(xlim = c(0, 100)) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(
    title = "Maths",
    x = "Maths Score"
  )

p2 <- ggplot(exam_data, aes(x = ENGLISH)) +
  geom_dotplot_interactive(
    aes(data_id = ID, tooltip = tooltip_text),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot"
  ) +
  coord_cartesian(xlim = c(0, 100)) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(
    title = "English",
    x = "English Score"
  )

girafe(
  code = print(p1 + p2),
  width_svg = 9,
  height_svg = 4,
  options = list(
    opts_tooltip(css = tooltip_css),
    opts_hover(css = "fill: #202020;"),
    opts_hover_inv(css = "opacity:0.2;")
  )
)
```

# 2. Interactive Data Visualisation with plotly

## 2.1 Basic Interactive Scatter Plot

```{r}
plot_ly(
  data = exam_data,
  x = ~MATHS,
  y = ~ENGLISH,
  type = "scatter",
  mode = "markers"
)
```

## 2.2 Scatter Plot Coloured by Race

```{r}
plot_ly(
  data = exam_data,
  x = ~ENGLISH,
  y = ~MATHS,
  color = ~RACE,
  type = "scatter",
  mode = "markers",
  text = ~paste(
    "ID:", ID,
    "<br>Class:", CLASS,
    "<br>Gender:", GENDER,
    "<br>Race:", RACE,
    "<br>English:", ENGLISH,
    "<br>Maths:", MATHS
  ),
  hoverinfo = "text"
)
```

## 2.3 ggplot Converted to Interactive Plotly

```{r}
p <- ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point(size = 1) +
  coord_cartesian(
    xlim = c(0, 100),
    ylim = c(0, 100)
  ) +
  labs(
    title = "English vs Maths",
    x = "Maths Score",
    y = "English Score"
  )

ggplotly(p)
```

## 2.4 Coordinated Multiple Views with plotly

```{r}
d <- highlight_key(exam_data)

p1 <- ggplot(d, aes(x = MATHS, y = ENGLISH)) +
  geom_point(size = 1) +
  coord_cartesian(
    xlim = c(0, 100),
    ylim = c(0, 100)
  ) +
  labs(
    title = "English vs Maths",
    x = "Maths",
    y = "English"
  )

p2 <- ggplot(d, aes(x = MATHS, y = SCIENCE)) +
  geom_point(size = 1) +
  coord_cartesian(
    xlim = c(0, 100),
    ylim = c(0, 100)
  ) +
  labs(
    title = "Science vs Maths",
    x = "Maths",
    y = "Science"
  )

subplot(
  ggplotly(p1),
  ggplotly(p2),
  titleX = TRUE,
  titleY = TRUE
)
```

# 3. Interactive Data Table with DT

```{r}
DT::datatable(
  exam_data,
  class = "compact stripe hover",
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    searchHighlight = TRUE
  )
)
```

# 4. Linked Brushing with crosstalk

```{r}
shared_exam <- highlight_key(exam_data)

p <- ggplot(shared_exam, aes(x = ENGLISH, y = MATHS)) +
  geom_point(size = 1) +
  coord_cartesian(
    xlim = c(0, 100),
    ylim = c(0, 100)
  ) +
  labs(
    title = "Linked Brushing: English vs Maths",
    x = "English Score",
    y = "Maths Score"
  )

linked_plot <- highlight(
  ggplotly(p),
  "plotly_selected"
)

crosstalk::bscols(
  linked_plot,
  DT::datatable(
    shared_exam,
    class = "compact",
    options = list(
      pageLength = 8,
      scrollX = TRUE
    )
  ),
  widths = 6
)
```

# Learning Reflection

In this exercise, I learned how interactive visualisation can improve exploratory data analysis. `ggiraph` is useful when I want to add tooltip, hover, and click behaviour to ggplot graphics. `plotly` is useful for quick interactive scatter plots and converting ggplot charts into web-ready visualisations. `DT` helps present raw data interactively, while `crosstalk` allows linked interactions between plots and tables.
