---
title: "Hands-on Exercise 9a"
subtitle: "Choropleth Mapping with R"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Overview

This hands-on exercise recreates the key techniques from Chapter 21 of *R for Visual Analytics*. A **choropleth map** shades geographic areas in proportion to a statistical value. Using Singapore's Master Plan 2014 subzone boundaries and 2020 population, this exercise builds choropleth maps of the age **dependency ratio** with the `sf` and `tmap` packages.

> **Note:** `sf` and `tmap` depend on the system libraries GDAL, GEOS and PROJ. If a map fails to render, install those libraries first (e.g. `brew install gdal` on macOS).

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(sf, tmap, tidyverse)
```

## Importing the Data

```{r}
#| message: false
#| warning: false
#| results: hide

mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")

popdata <- read_csv("data/aspatial/respop2020.csv")
```

## Data Preparation

The 2020 population is summarised per subzone and grouped into three age bands - **young** (0-24), **economically active** (25-64) and **aged** (65+) - then the dependency ratio is computed.

```{r}
#| message: false
#| warning: false

young_ag  <- c("0_to_4", "5_to_9", "10_to_14", "15_to_19", "20_to_24")
active_ag <- c("25_to_29", "30_to_34", "35_to_39", "40_to_44",
               "45_to_49", "50_to_54", "55_to_59", "60_to_64")
aged_ag   <- c("65_to_69", "70_to_74", "75_to_79", "80_to_84",
               "85_to_89", "90_and_over")

popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  mutate(band = case_when(
    AG %in% young_ag  ~ "YOUNG",
    AG %in% active_ag ~ "ECONOMY ACTIVE",
    AG %in% aged_ag   ~ "AGED"
  )) %>%
  group_by(PA, SZ, band) %>%
  summarise(POP = sum(Pop), .groups = "drop") %>%
  pivot_wider(names_from = band, values_from = POP, values_fill = 0) %>%
  mutate(
    TOTAL = YOUNG + `ECONOMY ACTIVE` + AGED,
    DEPENDENCY = (YOUNG + AGED) / `ECONOMY ACTIVE`
  ) %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

The attribute table is joined onto the subzone geometry using the subzone name.

```{r}
mpsz_pop2020 <- left_join(
  mpsz, popdata2020,
  by = c("SUBZONE_N" = "SZ")
)
```

# 1. Quick Choropleth with `qtm()`

```{r}
#| fig-width: 9
#| fig-height: 6
tmap_mode("plot")
qtm(mpsz_pop2020, fill = "DEPENDENCY")
```

# 2. A Fully Customised Choropleth

```{r}
#| fig-width: 9
#| fig-height: 6
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(style = "quantile", values = "brewer.blues"),
    fill.legend = tm_legend(title = "Dependency ratio"),
    col = "grey40",
    lwd = 0.4
  ) +
  tm_title("Age Dependency Ratio by Subzone, 2020") +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(col_alpha = 0.2) +
  tm_layout(frame = TRUE)
```

# 3. Comparing Classification Methods

Different data-classification methods can tell very different visual stories from the same numbers.

```{r}
#| fig-width: 11
#| fig-height: 9
cmap <- function(style, ttl) {
  tm_shape(mpsz_pop2020) +
    tm_polygons(
      fill = "DEPENDENCY",
      fill.scale = tm_scale_intervals(n = 6, style = style, values = "brewer.blues")
    ) +
    tm_title(ttl)
}

tmap_arrange(
  cmap("equal", "Equal interval"),
  cmap("quantile", "Quantile"),
  cmap("jenks", "Natural breaks (jenks)"),
  cmap("kmeans", "k-means"),
  ncol = 2
)
```

## 3.1 Custom Breaks

Rather than an algorithm, the breaks can be set by hand from the data's own summary statistics.

```{r}
#| fig-width: 9
#| fig-height: 6
summary(mpsz_pop2020$DEPENDENCY)

tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00),
      values = "brewer.blues"
    )
  ) +
  tm_title("Custom breaks")
```

# Learning Reflection

In this exercise, I learned that a choropleth map is really two datasets joined: a **geometry** (the subzone boundaries in the shapefile) and an **attribute** (the population statistic), linked by a common key. Most of the effort is in the attribute preparation - aggregating age bands and computing the dependency ratio - before `tmap` can shade the polygons. I also learned that the **classification method** (`quantile` vs `equal` interval) strongly changes the story a choropleth tells, so it must be chosen deliberately rather than left at the default.
