---
title: "Hands-on Exercise 5a"
subtitle: "Visual Correlation Analysis"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Overview

This hands-on exercise recreates the key techniques from Chapter 6 of *R for Visual Analytics*. The focus is on building **correlation matrices and corrgrams** to reveal the pairwise relationships among the physicochemical properties of wine, using `corrplot` and `ggstatsplot`.

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(
  corrplot,
  ggstatsplot,
  tidyverse
)

wine <- read_csv("data/wine_quality.csv")
```

## Data Preview

```{r}
DT::datatable(
  head(wine, 50),
  class = "compact",
  options = list(pageLength = 10, scrollX = TRUE)
)
```

# 1. Building a Correlation Matrix with `pairs()`

## 1.1 Basic Scatterplot Matrix

```{r}
pairs(wine[, 1:11])
```

## 1.2 Lower-corner Only

Showing only the lower half removes the redundant mirror image.

```{r}
pairs(wine[, 2:12], upper.panel = NULL)
```

## 1.3 Showing the Correlation Coefficient

A custom `panel.cor()` function prints the absolute correlation in the upper panel, sized by strength.

```{r}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...) {
  usr <- par("usr")
  on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y, use = "complete.obs"))
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  if (missing(cex.cor)) cex.cor <- 0.8 / strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * (1 + r) / 2)
}

pairs(wine[, 2:12], upper.panel = panel.cor)
```

# 2. Corrgram with ggstatsplot

## 2.1 `ggcorrmat()`

`ggstatsplot::ggcorrmat()` draws the correlation coefficients and marks the statistically non-significant pairs with a cross.

```{r}
#| fig-width: 8
#| fig-height: 7
ggstatsplot::ggcorrmat(
  data = wine,
  cor.vars = 1:11
)
```

## 2.2 Grouped Corrgram by Wine Type

```{r}
#| fig-width: 11
#| fig-height: 6
grouped_ggcorrmat(
  data = wine,
  cor.vars = 1:11,
  grouping.var = type,
  plotgrid.args = list(ncol = 2)
)
```

# 3. Corrgram with corrplot

## 3.1 Computing and Plotting the Correlation Matrix

```{r}
wine.cor <- cor(wine[, 1:11])
corrplot(wine.cor)
```

## 3.2 Changing the Visual Element

Using ellipses and showing only the lower half.

```{r}
corrplot(
  wine.cor,
  method = "ellipse",
  type = "lower"
)
```

## 3.3 Mixed Layout

Ellipses in the lower half, coefficients in the upper half.

```{r}
corrplot.mixed(
  wine.cor,
  lower = "ellipse",
  upper = "number",
  tl.pos = "lt",
  diag = "l",
  tl.col = "black"
)
```

## 3.4 Combining with a Significance Test

Pairs whose correlation is not significant at the 0.05 level are left blank.

```{r}
wine.sig <- cor.mtest(wine[, 1:11], conf.level = 0.95)

corrplot(
  wine.cor,
  method = "number",
  type = "lower",
  diag = FALSE,
  tl.col = "black",
  tl.srt = 45,
  p.mat = wine.sig$p,
  sig.level = 0.05
)
```

## 3.5 Reordering with Hierarchical Clustering

`corrplot` can reorder the variables so that strongly correlated groups sit together, with cluster rectangles drawn on top.

```{r}
corrplot(
  wine.cor,
  method = "ellipse",
  tl.pos = "lt",
  tl.col = "black",
  order = "hclust",
  hclust.method = "ward.D",
  addrect = 3
)
```

# Learning Reflection

In this exercise, I learned that a correlation matrix is far easier to read as a **corrgram** than as a table of numbers. `pairs()` gives a quick raw look, but `ggstatsplot::ggcorrmat()` adds real value by flagging non-significant relationships, and `grouped_ggcorrmat()` lets me compare red and white wine side by side. `corrplot` is the most flexible: reordering by hierarchical clustering and drawing cluster rectangles turned a noisy 11×11 matrix into clear blocks of related chemical properties.
