---
title: "Hands-on Exercise 3b"
subtitle: "Animated Statistical Graphics"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

# Part 2: Programming Animated Statistical Graphics with R

This section extends Hands-on Exercise 3 by creating animated statistical graphics using `gganimate` and `plotly`.

## Setup for Animated Graphics

```{r}
#| message: false
#| warning: false

pacman::p_load(
  readxl,
  gifski,
  gapminder,
  plotly,
  gganimate,
  tidyverse
)
```

## Import Global Population Data

```{r}
globalPop <- read_xls("GlobalPopulation.xls", sheet = "Data") %>%
  mutate(
    across(c(Country, Continent), as.factor),
    Year = as.integer(Year)
  )
```

## Static Population Bubble Plot

```{r}
ggplot(globalPop, aes(
  x = Old,
  y = Young,
  size = Population,
  colour = Country
)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = gapminder::country_colors) +
  scale_size(range = c(2, 12)) +
  labs(
    title = "Population Structure",
    x = "% Aged",
    y = "% Young"
  )
```

## Animated Bubble Plot using gganimate

```{r}
#| fig-width: 8
#| fig-height: 5

p <- ggplot(globalPop, aes(
  x = Old,
  y = Young,
  size = Population,
  colour = Country
)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = gapminder::country_colors) +
  scale_size(range = c(2, 12)) +
  labs(
    title = "Year: {frame_time}",
    x = "% Aged",
    y = "% Young"
  ) +
  transition_time(Year) +
  ease_aes("linear")

animate(p, renderer = gifski_renderer())
```

## Animated Bubble Plot using ggplotly

```{r}
gg <- ggplot(globalPop, aes(
  x = Old,
  y = Young,
  size = Population,
  colour = Country
)) +
  geom_point(
    aes(frame = Year),
    alpha = 0.7
  ) +
  scale_colour_manual(values = gapminder::country_colors) +
  scale_size(range = c(2, 12)) +
  labs(
    x = "% Aged",
    y = "% Young"
  ) +
  theme(legend.position = "none")

ggplotly(gg)
```

## Animated Bubble Plot using plot_ly

```{r}
globalPop %>%
  plot_ly(
    x = ~Old,
    y = ~Young,
    size = ~Population,
    color = ~Continent,
    sizes = c(2, 100),
    frame = ~Year,
    text = ~Country,
    hoverinfo = "text",
    type = "scatter",
    mode = "markers"
  ) %>%
  layout(showlegend = FALSE)
```

## Learning Reflection

In this section, I learned how animation can be used to show changes over time. `gganimate` creates GIF-style animations from ggplot objects, while `plotly` creates interactive animated charts with play buttons and sliders.
