---
title: "Hands-on Exercise 4b"
subtitle: "Visual Statistical Analysis"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    code-fold: true
---

## Overview

This hands-on exercise focuses on visual statistical analysis using the `ggstatsplot` package.

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(
  ggstatsplot,
  tidyverse
)

exam <- read_csv("Exam_data.csv")
```

## Data Preview

```{r}
glimpse(exam)
```

# 1. One-sample Test

## 1.1 One-sample Test for English Scores

```{r}
set.seed(1234)

gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores"
)
```

# 2. Two-sample Mean Test

## 2.1 Maths Scores by Gender

```{r}
ggbetweenstats(
  data = exam,
  x = GENDER,
  y = MATHS,
  type = "np",
  messages = FALSE
)
```

# 3. One-way ANOVA Test

## 3.1 English Scores by Race

```{r}
ggbetweenstats(
  data = exam,
  x = RACE,
  y = ENGLISH,
  type = "p",
  mean.ci = TRUE,
  pairwise.comparisons = TRUE,
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```

# 4. Correlation Test

## 4.1 Maths Scores vs English Scores

```{r}
ggscatterstats(
  data = exam,
  x = MATHS,
  y = ENGLISH,
  marginal = FALSE
)
```

# 5. Association Test

## 5.1 Binning Maths Scores

```{r}
exam1 <- exam %>%
  mutate(
    MATHS_bins = cut(
      MATHS,
      breaks = c(0, 60, 75, 85, 100),
      include.lowest = TRUE,
      labels = c("0-60", "61-75", "76-85", "86-100")
    )
  )

exam1 %>%
  count(MATHS_bins, GENDER)
```

## 5.2 Association Between Maths Score Bands and Gender

```{r}
ggbarstats(
  data = exam1,
  x = MATHS_bins,
  y = GENDER
)
```

# 6. Additional Exploration

## 6.1 Science Scores by Gender

```{r}
ggbetweenstats(
  data = exam,
  x = GENDER,
  y = SCIENCE,
  type = "np",
  messages = FALSE
)
```

## 6.2 Science Scores vs Maths Scores

```{r}
ggscatterstats(
  data = exam,
  x = SCIENCE,
  y = MATHS,
  marginal = FALSE
)
```

## Learning Reflection

In this exercise, I learned that `ggstatsplot` is useful because it combines statistical test results with visualisation. Instead of only showing the chart, it also displays useful statistical information such as test statistics, p-values, confidence intervals, and sample sizes.
