---
title: "Hands-on Exercise 5e"
subtitle: "Treemap Visualisation with R"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Overview

This hands-on exercise recreates the key techniques from Chapter 16 of *R for Visual Analytics*. A **treemap** displays hierarchical, part-to-whole data as nested rectangles, where area encodes one measure and colour encodes another. Using 2018 Singapore private residential transactions, this exercise builds treemaps with the `treemap` and `treemapify` packages.

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(treemap, treemapify, d3treeR, tidyverse)

realis2018 <- read_csv("data/realis2018.csv")
```

## Data Preparation

Transactions are grouped by planning region, planning area and project. For each project we sum the units sold and take the median unit price.

```{r}
#| message: false
#| warning: false

realis2018_grouped <- realis2018 %>%
  group_by(`Planning Region`, `Planning Area`, `Project Name`) %>%
  summarise(
    `Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE),
    `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
    `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
    `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
DT::datatable(
  realis2018_grouped,
  class = "compact",
  options = list(pageLength = 10, scrollX = TRUE)
)
```

# 1. Treemap with the treemap Package

## 1.1 Basic Treemap

Rectangle **size** = total units sold; **colour** = median unit price. The hierarchy is region → planning area → project.

```{r}
#| fig-width: 9
#| fig-height: 7
treemap(
  realis2018_grouped,
  index = c("Planning Region", "Planning Area", "Project Name"),
  vSize = "Total Unit Sold",
  vColor = "Median Unit Price ($ psm)",
  type = "value",
  title = "Private Residential Property Sold, 2018",
  title.legend = "Median Unit Price (S$ psm)"
)
```

## 1.2 Selecting a Single Region

Focusing on the Central Region reduces clutter and makes the individual projects legible.

```{r}
#| fig-width: 9
#| fig-height: 7
realis_central <- realis2018_grouped %>%
  filter(`Planning Region` == "Central Region")

treemap(
  realis_central,
  index = c("Planning Area", "Project Name"),
  vSize = "Total Unit Sold",
  vColor = "Median Unit Price ($ psm)",
  type = "value",
  title = "Central Region Private Residential Property Sold, 2018",
  title.legend = "Median Unit Price (S$ psm)"
)
```

# 2. Treemap with treemapify (ggplot2)

`treemapify` provides `geom_treemap`, letting the treemap live inside the familiar ggplot2 grammar.

```{r}
#| fig-width: 9
#| fig-height: 6
realis_area <- realis2018_grouped %>%
  group_by(`Planning Region`, `Planning Area`) %>%
  summarise(
    `Total Unit Sold` = sum(`Total Unit Sold`, na.rm = TRUE),
    `Median Unit Price ($ psm)` = median(`Median Unit Price ($ psm)`, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(
  realis_area,
  aes(
    area = `Total Unit Sold`,
    fill = `Median Unit Price ($ psm)`,
    label = `Planning Area`,
    subgroup = `Planning Region`
  )
) +
  geom_treemap() +
  geom_treemap_subgroup_border(colour = "white") +
  geom_treemap_text(colour = "white", place = "centre", grow = FALSE, reflow = TRUE) +
  scale_fill_gradient(low = "#deebf7", high = "#08519c") +
  labs(title = "Units Sold by Planning Area, 2018")
```

# 3. Interactive Treemap with d3treeR

`d3treeR::d3tree()` turns a `treemap` object into an interactive, drill-down visualisation - click a rectangle to zoom into its children. We first build a `treemap` object and save it to `tm`, then pass it to `d3tree()`.

```{r}
#| fig-width: 9
#| fig-height: 7
tm <- treemap(
  realis2018_grouped,
  index = c("Planning Region", "Planning Area"),
  vSize = "Total Unit Sold",
  vColor = "Median Unit Price ($ psm)",
  type = "value",
  title = "Private Residential Property Sold, 2018",
  title.legend = "Median Unit Price (S$ psm)"
)

d3tree(tm, rootname = "Singapore")
```

# Learning Reflection

In this exercise, I learned that a treemap is well suited to hierarchical, part-to-whole data because it uses both **area** and **colour** to carry two measures at once - here, units sold and median price. As with the other multivariate exercises, most of the work is in the data preparation: grouping to the right level of the hierarchy before plotting. The `treemap` package gives a fast, information-dense result, while `treemapify` integrates with ggplot2 so I can theme it and combine it with the rest of my visual grammar.
