---
title: "Hands-on Exercise 1a"
subtitle: "A Layered Grammar of Graphics: ggplot2 methods"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Overview

This hands-on exercise recreates the key techniques from Chapter 1 of *R for Visual Analytics*. The focus is on the **Layered Grammar of Graphics** - how `ggplot2` builds a plot from independent building blocks: data, aesthetic mappings, geometric objects, statistics, facets, coordinates, and themes.

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(tidyverse)

exam_data <- read_csv("Exam_data.csv")
```

## Data Preview

```{r}
DT::datatable(
  exam_data,
  class = "compact",
  options = list(
    pageLength = 10,
    scrollX = TRUE
  )
)
```

# 1. Essential Grammatical Elements: Data and Aesthetics

## 1.1 The `data` and `aes()` Layers

A plot begins with a call to `ggplot()`, which initialises a coordinate system using the supplied data and aesthetic mappings. On its own it produces a blank canvas.

```{r}
ggplot(data = exam_data, aes(x = MATHS))
```

# 2. Geometric Objects

## 2.1 Bar Chart with `geom_bar()`

```{r}
ggplot(exam_data, aes(x = RACE)) +
  geom_bar()
```

## 2.2 Dot Plot with `geom_dotplot()`

`scale_y_continuous()` is used to turn off the misleading y-axis.

```{r}
ggplot(exam_data, aes(x = MATHS)) +
  geom_dotplot(binwidth = 2.5, dotsize = 0.5) +
  scale_y_continuous(NULL, breaks = NULL)
```

## 2.3 Histogram with `geom_histogram()`

```{r}
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue")
```

## 2.4 Histogram Coloured by a Sub-group

```{r}
ggplot(exam_data, aes(x = MATHS, fill = GENDER)) +
  geom_histogram(bins = 20, color = "grey30")
```

## 2.5 Kernel Density Estimate with `geom_density()`

```{r}
ggplot(exam_data, aes(x = MATHS, colour = GENDER)) +
  geom_density()
```

## 2.6 Boxplot and Notched Boxplot with `geom_boxplot()`

```{r}
ggplot(exam_data, aes(y = MATHS, x = GENDER)) +
  geom_boxplot(notch = TRUE)
```

## 2.7 Violin Plot with `geom_violin()`

```{r}
ggplot(exam_data, aes(y = MATHS, x = GENDER)) +
  geom_violin()
```

## 2.8 Scatterplot with `geom_point()`

```{r}
ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point()
```

## 2.9 Combining Geometric Objects

Boxplot and scatter (jittered points) can be layered in a single plot.

```{r}
ggplot(exam_data, aes(y = MATHS, x = GENDER)) +
  geom_boxplot() +
  geom_point(position = "jitter", size = 0.5)
```

# 3. Statistics

## 3.1 Adding a Mean with `stat_summary()`

```{r}
ggplot(exam_data, aes(y = MATHS, x = GENDER)) +
  geom_boxplot() +
  stat_summary(
    geom = "point",
    fun = "mean",
    colour = "red",
    size = 4
  )
```

## 3.2 Adding a Best-fit Curve with `geom_smooth()`

```{r}
ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm, linewidth = 0.5)
```

# 4. Facets

## 4.1 `facet_wrap()`

```{r}
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20) +
  facet_wrap(~ CLASS)
```

## 4.2 `facet_grid()`

```{r}
ggplot(exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 20) +
  facet_grid(GENDER ~ CLASS)
```

# 5. Coordinates

## 5.1 Flipping Axes with `coord_flip()`

```{r}
ggplot(exam_data, aes(x = RACE)) +
  geom_bar() +
  coord_flip()
```

## 5.2 Fixing the Range with `coord_cartesian()`

```{r}
ggplot(exam_data, aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm, linewidth = 0.5) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100))
```

# 6. Themes

## 6.1 Comparing Built-in Themes

```{r}
ggplot(exam_data, aes(x = RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray() +
  labs(title = "theme_gray()")
```

```{r}
ggplot(exam_data, aes(x = RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic() +
  labs(title = "theme_classic()")
```

```{r}
ggplot(exam_data, aes(x = RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal() +
  labs(title = "theme_minimal()")
```

# Learning Reflection

In this exercise, I learned that the Layered Grammar of Graphics treats a chart as a set of independent building blocks - data, aesthetic mappings, geometric objects, statistics, facets, coordinates, and themes - that stack together with the `+` operator. Because each layer is independent, I can start from the same `ggplot()` call and swap the `geom`, add a `stat`, split with a `facet`, or restyle with a `theme` without rebuilding the plot from scratch. This modular thinking is the foundation for every later exercise.
