---
title: "Hands-on Exercise 4c"
subtitle: "Visualising Uncertainty"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    code-fold: true
---

## Overview

This hands-on exercise focuses on visualising uncertainty using error bars, confidence intervals, interactive charts, and uncertainty visualisation methods.

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(
  plotly,
  crosstalk,
  DT,
  ggdist,
  ggridges,
  colorspace,
  gganimate,
  tidyverse,
  knitr
)

exam <- read_csv("Exam_data.csv")
```

## Data Preview

```{r}
glimpse(exam)
```

# 1. Summary Statistics

```{r}
my_sum <- exam %>%
  group_by(RACE) %>%
  summarise(
    n = n(),
    mean = mean(MATHS, na.rm = TRUE),
    sd = sd(MATHS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    se = sd / sqrt(n - 1)
  )

knitr::kable(my_sum, format = "html", digits = 2)
```

# 2. Visualising Uncertainty with Error Bars

## 2.1 Standard Error Bars

```{r}
ggplot(my_sum) +
  geom_errorbar(
    aes(
      x = RACE,
      ymin = mean - se,
      ymax = mean + se
    ),
    width = 0.2,
    colour = "black",
    alpha = 0.9,
    linewidth = 0.5
  ) +
  geom_point(
    aes(
      x = RACE,
      y = mean
    ),
    stat = "identity",
    color = "red",
    size = 1.5,
    alpha = 1
  ) +
  labs(
    title = "Standard Error of Mean Maths Score by Race",
    x = "Race",
    y = "Mean Maths score"
  ) +
  theme_minimal()
```

## 2.2 95% Confidence Interval

```{r}
ggplot(my_sum) +
  geom_errorbar(
    aes(
      x = reorder(RACE, -mean),
      ymin = mean - 1.96 * se,
      ymax = mean + 1.96 * se
    ),
    width = 0.2,
    colour = "black",
    alpha = 0.9,
    linewidth = 0.5
  ) +
  geom_point(
    aes(
      x = reorder(RACE, -mean),
      y = mean
    ),
    stat = "identity",
    color = "red",
    size = 1.5,
    alpha = 1
  ) +
  labs(
    title = "95% Confidence Interval of Mean Maths Score by Race",
    x = "Race",
    y = "Mean Maths score"
  ) +
  theme_minimal()
```

# 3. Interactive Error Bars

## 3.1 99% Confidence Interval with Linked Table

```{r}
shared_df <- SharedData$new(my_sum)

p <- ggplot(shared_df) +
  geom_errorbar(
    aes(
      x = reorder(RACE, -mean),
      ymin = mean - 2.58 * se,
      ymax = mean + 2.58 * se
    ),
    width = 0.2,
    colour = "black",
    alpha = 0.9,
    linewidth = 0.5
  ) +
  geom_point(
    aes(
      x = RACE,
      y = mean,
      text = paste(
        "Race:", RACE,
        "<br>N:", n,
        "<br>Average score:", round(mean, 2),
        "<br>99% CI: [",
        round(mean - 2.58 * se, 2),
        ",",
        round(mean + 2.58 * se, 2),
        "]"
      )
    ),
    stat = "identity",
    color = "red",
    size = 1.5,
    alpha = 1
  ) +
  xlab("Race") +
  ylab("Average Maths Score") +
  ggtitle("99% Confidence Interval of Average Maths Scores by Race") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      vjust = 0.5,
      hjust = 1
    )
  )

bscols(
  widths = c(6, 6),
  ggplotly(p, tooltip = "text"),
  DT::datatable(
    shared_df,
    rownames = FALSE,
    class = "compact",
    width = "100%",
    options = list(
      pageLength = 10,
      scrollX = TRUE
    )
  ) %>%
    formatRound(
      columns = c("mean", "sd", "se"),
      digits = 2
    )
)
```

# 4. Visualising Uncertainty with ggdist

## 4.1 Mean Point Interval Plot

```{r}
exam %>%
  ggplot(
    aes(
      x = RACE,
      y = MATHS
    )
  ) +
  stat_pointinterval() +
  labs(
    title = "Visualising Confidence Intervals of Mean Maths Score",
    subtitle = "Mean point + multiple interval plot",
    x = "Race",
    y = "Maths score"
  ) +
  theme_minimal()
```

## 4.2 Median Point Interval Plot

```{r}
exam %>%
  ggplot(
    aes(
      x = RACE,
      y = MATHS
    )
  ) +
  stat_pointinterval(
    .width = 0.95,
    .point = median,
    .interval = qi
  ) +
  labs(
    title = "Visualising Confidence Intervals of Median Maths Score",
    subtitle = "Median point + interval plot",
    x = "Race",
    y = "Maths score"
  ) +
  theme_minimal()
```

## 4.3 Multiple Interval Plot

```{r}
exam %>%
  ggplot(
    aes(
      x = RACE,
      y = MATHS
    )
  ) +
  stat_pointinterval(
    .width = c(0.95, 0.99),
    show.legend = TRUE
  ) +
  labs(
    title = "Visualising 95% and 99% Intervals of Mean Maths Score",
    subtitle = "Mean point + multiple interval plot",
    x = "Race",
    y = "Maths score"
  ) +
  theme_minimal()
```

## 4.4 Gradient Interval Plot

```{r}
exam %>%
  ggplot(
    aes(
      x = RACE,
      y = MATHS
    )
  ) +
  stat_gradientinterval(
    fill = "skyblue",
    show.legend = TRUE
  ) +
  labs(
    title = "Visualising Confidence Intervals of Mean Maths Score",
    subtitle = "Gradient + interval plot",
    x = "Race",
    y = "Maths score"
  ) +
  theme_minimal()
```

# 5. Optional: Hypothetical Outcome Plot

This section is optional because the `ungeviz` package may need to be installed from GitHub. To avoid rendering errors, the code is shown but not executed.

```{r}
#| eval: false

# install.packages("devtools")
# devtools::install_github("wilkelab/ungeviz")

library(ungeviz)

ggplot(
  data = exam,
  aes(
    x = factor(RACE),
    y = MATHS
  )
) +
  geom_point(
    position = position_jitter(
      height = 0.3,
      width = 0.05
    ),
    size = 0.4,
    color = "#0072B2",
    alpha = 1 / 2
  ) +
  geom_hpline(
    data = sampler(25, group = RACE),
    height = 0.6,
    color = "#D55E00"
  ) +
  theme_bw() +
  transition_states(.draw, 1, 3)
```

## Learning Reflection

In this exercise, I learned that uncertainty should not be hidden in visual analytics. Error bars, confidence intervals, and interval plots help communicate how reliable or uncertain a point estimate is. Interactive visualisation also helps users inspect the uncertainty values more clearly.
