---
title: "Hands-on Exercise 5d"
subtitle: "Visual Multivariate Analysis with Parallel Coordinates Plot"
author: "Jedidiah Asaf T"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
---

## Overview

This hands-on exercise recreates the key techniques from Chapter 15 of *R for Visual Analytics*. A **parallel coordinates plot** draws one vertical axis per variable and connects each observation as a line, making it ideal for comparing many quantitative indicators at once. Using the 2018 World Happiness data, this exercise builds static plots with `GGally` and an interactive one with `parallelPlot`.

## Setup

```{r}
#| message: false
#| warning: false

if (!require(pacman)) install.packages("pacman")

pacman::p_load(GGally, parallelPlot, tidyverse)

wh <- read_csv("data/WHData-2018.csv")
```

## Data Preview

```{r}
DT::datatable(
  wh,
  class = "compact",
  options = list(pageLength = 10, scrollX = TRUE)
)
```

# 1. Static Parallel Coordinates with GGally

## 1.1 Basic Plot

Columns 7-12 are the six contributing factors of the happiness score.

```{r}
#| fig-width: 9
#| fig-height: 5
ggparcoord(
  data = wh,
  columns = c(7:12)
)
```

## 1.2 With Grouping, Scaling and Boxplot

Colouring by `Region`, scaling each axis with `uniminmax`, and overlaying a boxplot makes the group differences readable.

```{r}
#| fig-width: 11
#| fig-height: 6
ggparcoord(
  data = wh,
  columns = c(7:12),
  groupColumn = 2,
  scale = "uniminmax",
  alphaLines = 0.2,
  boxplot = TRUE,
  title = "Parallel Coordinates Plot of World Happiness Variables"
)
```

## 1.3 Faceting by Region

```{r}
#| fig-width: 12
#| fig-height: 8
ggparcoord(
  data = wh,
  columns = c(7:12),
  groupColumn = 2,
  scale = "uniminmax",
  alphaLines = 0.2,
  boxplot = TRUE,
  title = "Parallel Coordinates Plots faceted by Region"
) +
  facet_wrap(~ Region) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 30, size = 7)
  )
```

# 2. Interactive Parallel Coordinates with parallelPlot

## 2.1 Basic Interactive Plot

```{r}
wh_sel <- wh %>%
  select(`Happiness score`, c(7:12))

parallelPlot(
  wh_sel,
  width = 320,
  height = 250
)
```

## 2.2 With Rotated Axis Titles and Histograms

```{r}
parallelPlot(
  wh_sel,
  rotateTitle = TRUE,
  histoVisibility = rep(TRUE, ncol(wh_sel))
)
```

# Learning Reflection

In this exercise, I learned that a parallel coordinates plot is the natural choice when I need to compare more than two continuous variables together. The key is **scaling** - using `uniminmax` puts every axis on a common 0-1 range so a variable with large values does not visually swamp the others. `GGally::ggparcoord()` with grouping, a boxplot overlay and region facets exposes the multivariate profile of each region, while `parallelPlot` makes it interactive, letting me brush lines and add per-axis histograms to inspect distributions.
